Developmental trajectories of word recognition: a priming study in bilinguals

### 2023-10-26 manuscript/manuscript.qmd
Add supplementary vocabulary checklists and describe GAMMS

```{r setup}
#| label: setup
#| echo: false
#| message: false
#| warning: false
# load objects -----------------------------------------------------------------

targets::tar_config_set(store = here::here('_targets'),
						script = here::here('_targets.R'))

@@

targets::tar_load(
	c(participants,
	  participants_oxf,
	  stimuli,
	  stimuli_oxf,
	  vocabulary,
	  vocabulary_oxf,
	  attrition_trials, 
	  attrition_participants,
	  attrition_trials_vtarget,
	  attrition_participants_vtarget,
	  attrition_trials_vnone, 
	  attrition_participants_vnone,
	  attrition_trials_noeach,
	  attrition_participants_noeach,
	  attrition_trials_vnoeach,
	  attrition_participants_vnoeach,
	  attrition_trials_oxf, 
	  attrition_participants_oxf,
	  attrition_trials_vtarget_oxf,
	  attrition_participants_vtarget_oxf,
	  attrition_trials_vnone_oxf, 
	  attrition_participants_vnone_oxf,
	  attrition_trials_noeach_oxf,
	  attrition_participants_noeach_oxf,
	  attrition_trials_vnoeach_oxf,
	  attrition_participants_vnoeach_oxf,
	  bvq_data, 
	  gaze_aoi, 
	  looking_times,
	  data_time_related, 
	  data_time_cognate,
	  data_time_related_vtarget,
	  data_time_cognate_vtarget,
	  data_time_related_vnone,
	  data_time_cognate_vnone,
	  data_time_related_noeach, 
	  data_time_cognate_noeach,
	  data_time_related_vnoeach,
	  data_time_cognate_vnoeach,
	  data_time_related_oxf, 
	  data_time_cognate_oxf,
	  data_time_related_vtarget_oxf,
	  data_time_cognate_vtarget_oxf,
	  data_time_related_vnone_oxf,
	  data_time_cognate_vnone_oxf,
	  data_time_related_noeach_oxf, 
	  data_time_cognate_noeach_oxf,
	  data_time_related_vnoeach_oxf,
	  data_time_cognate_vnoeach_oxf
	))

targets::tar_load(
	c(model_fits_related,
	  model_fits_cognate,
	  model_fits_related_vtarget,
	  model_fits_cognate_vtarget,
	  model_fits_related_vnone,
	  model_fits_cognate_vnone,
	  model_fits_related_noeach,
	  model_fits_cognate_noeach,
	  model_fits_related_vnoeach,
	  model_fits_cognate_vnoeach
	  # model_loo_related,
	  # model_loo_cognate,
	  # model_loo_related_vtarget,
	  # model_loo_cognate_vtarget,
	  # model_loo_related_vnone,
	  # model_loo_cognate_vnone,
	  # model_loo_related_noeach,
	  # model_loo_cognate_noeach,
	  # model_loo_related_vnoeach,
	  # model_loo_cognate_vnoeach
	))

library(knitr)
library(kableExtra)
library(ggplot2)
library(gt)
library(patchwork)
library(english)

# set ggplot theme and colour palette ------------------------------------------

@@

# prepare data
attrition <- lst(participants,
				 attrition_participants,
				 attrition_trials) |> 
	purrr::reduce(left_join, by = join_by(filename))

attrition_vtarget <- lst(participants,
						 attrition_participants_vtarget,
						 attrition_trials_vtarget) |> 
	purrr::reduce(left_join, by = join_by(filename))

attrition_vnone <- lst(participants,
					   attrition_participants_vnone,
					   attrition_trials_vnone) |> 
	purrr::reduce(left_join, by = join_by(filename))

attrition_noeach <- lst(participants,
						attrition_participants_noeach,
						attrition_trials_noeach) |> 
	purrr::reduce(left_join, by = join_by(filename))

attrition_vnoeach <- lst(participants,
						 attrition_participants_vnoeach,
						 attrition_trials_vnoeach) |> 
	purrr::reduce(left_join, by = join_by(filename))

attrition_oxf <- lst(participants_oxf,
					 attrition_participants_oxf,
					 attrition_trials_oxf) |> 
	purrr::reduce(left_join, by = join_by(id))

attrition_vtarget_oxf <- lst(participants_oxf,
							 attrition_participants_vtarget_oxf,
							 attrition_trials_vtarget_oxf) |> 
	purrr::reduce(left_join, by = join_by(id))

attrition_vnone_oxf <- lst(participants_oxf,
						   attrition_participants_vnone_oxf,
						   attrition_trials_vnone_oxf) |> 
	purrr::reduce(left_join, by = join_by(id))

attrition_noeach_oxf <- lst(participants_oxf,
							attrition_participants_noeach_oxf,
							attrition_trials_noeach_oxf) |> 
	purrr::reduce(left_join, by = join_by(id))

attrition_vnoeach_oxf <- lst(participants_oxf,
							 attrition_participants_vnoeach_oxf,
							 attrition_trials_vnoeach_oxf) |> 
	purrr::reduce(left_join, by = join_by(id))

@@

```{r}
#| label: participants-numbers
n_participants_total <- length(unique(participants$id))

n_participants_sessions <- count(participants, id, name = "n_sessions") |> 
	count(n_sessions) |> 
	group_split(n_sessions) |> 
	set_names(paste0("session_", 1:3))

@@

n_sessions_age_group <- participants |> 
	summarise(across(age, lst(mean, sd, min, max)),
			  n = n(),
			  .by = c(age_group)) |> 
	mutate(across(age_mean:age_max, \(x) round(x, 2))) |> 
	group_split(age_group) |> 
	set_names(c("age_21", "age_25", "age_30"))

n_sessions_dominance <- count(participants, test_language) |> 
	group_split(test_language) |> 
	set_names(c("catalan", "spanish"))

n_sessions_dominance_age_group <- count(participants, age_group, test_language) |> 
	group_split(test_language) |> 
	set_names(c("catalan", "spanish")) |> 
	map(\(x) group_split(x, age_group) |> 
			set_names(c("age_21", "age_25", "age_30")))

n_sessions_lp <- participants |> 
	count(lp) |> 
	group_split(lp) |> 
	set_names(c("monolingual", "bilingual"))

n_sessions_lp_age_group <- participants |> 
	count(lp, age_group) |> 
	group_split(lp) |> 
	set_names(c("monolingual", "bilingual")) |> 
	map(\(x) group_split(x, age_group) |> 
			set_names(c("age_21", "age_25", "age_30")))
```

We collected data from `r n_participants_total` monolingual and bilingual participants living in the Metropolitan Area of Barcelona (Spain), who were exposed to at least Catalan and/or Spanish from birth. Families were recruited from maternity room in private hospitals in Barcelona, and contacted via phone when the child's age spanned between our age intervals of interest. Families were invited to participate at three age points: 21, 25, and 30 months. `r n_participants_sessions[[1]]$n` participants were tested at one age point,  `r n_participants_sessions[[2]]$n` at two age points, and `r n_participants_sessions[[3]]$n` at the three age points. In total, we gathered data from `r n_sessions_total$n` testing sessions: `r n_sessions_age_group$age_21$n` at 21 months (*Mean* = `r round(n_sessions_age_group$age_21$age_mean, 2)`, *SD* = `r round(n_sessions_age_group$age_21$age_sd, 2)`, *Range* = `r n_sessions_age_group$age_21$age_min`--`r n_sessions_age_group$age_21$age_max`), `r n_sessions_age_group$age_25$n` at 25 months (*Mean* = `r round(n_sessions_age_group$age_25$age_mean, 2)`, *SD* = `r round(n_sessions_age_group$age_25$age_sd, 2)`, *Range* = `r n_sessions_age_group$age_25$age_min`--`r n_sessions_age_group$age_25$age_max`), and `r n_sessions_age_group$age_30$n` at 30 months (*Mean* = `r round(n_sessions_age_group$age_30$age_mean, 2)`, *SD* = `r round(n_sessions_age_group$age_25$age_sd, 2)`, *Range* = `r n_sessions_age_group$age_30$age_min`--`r n_sessions_age_group$age_30$age_max`).

### Language profile {#sec-lp}

We assessed participants' language profile using the Language Exposure Questionnaire [LEQ, @bosch2001evidence]. Before each experimental session, the experimenter asked the caretakers to estimate the amount of hours per day they and other people in the infant's social circle have spent speaking to the infant in any language since birth. The output of this interview is an estimated degree of exposure (DoE) to each language, indicated by the proportion of time the infant was reported to have listened to each language. According to this estimate, we classified participants as Catalan- or Spanish-dominant if the language with highest DoE was Catalan or Spanish, respectively, and tested the participant in the stimuli set that contained words in their native language. We collected data from `r n_sessions_dominance$catalan$n` Catalan-dominant participants in Catalan (`r n_sessions_dominance_age_group$catalan$age_21$n` at 21 months, `r n_sessions_dominance_age_group$catalan$age_25$n` at 25 months, and `r n_sessions_dominance_age_group$catalan$age_30$n` at 30 months). We further classified participants as monolinguals if the DoE to their dominant language exceeded 80% of the total DoE to Catalan and Spanish, and as bilinguals otherwise. Participants with DoE to language other than Catalan or Spanish were excluded from analyses. This divided the sample into `r n_sessions_lp$monolingual$n` monolinguals (`r n_sessions_lp_age_group$monolinguals$age_21` at 21 months, `r n_sessions_lp_age_group$monolinguals$age_25` at 25 months, and `r n_sessions_lp_age_group$monolinguals$age_30` at 30 months), and `r n_sessions_lp$bilingual$n` bilinguals (`r n_sessions_lp_age_group$monolinguals$age_21` at 21 months, `r n_sessions_lp_age_group$bilingual$age_25` at 25 months, and `r n_sessions_lp_age_group$bilingual$age_30` at 30 months). @tbl-participants-lp shows a detailed description of the linguistic profile of our sample.

@@

```{r tbl-vocabulary}
#| label: tbl-vocabulary
#| tbl-cap: "Summary of participant vocabulary sizes. vocabulary sizes are expressed as percentage points, and summarised as the mean (*M*) and the standard deviation (*SD*). Total: proportion of words in both languages marked as *Understands*. L1: proportion of words in the dominant language marked as *Understands*. L2: proportion of words in the non-cominant language marked as *Understands*. Conceptual: proportion of translation equivalents for which at least one of the words has ben marked as *Understands*. TE: proportion of translation equivalents for which *both* words have been marked as *Understands*."
#| eval: false
vocabulary |> 
	left_join(select(participants, filename, id, age_group, lp),
			  by = join_by(filename)) |> 
	summarise(across(matches("prop"), 
					 tibble::lst(mean, sd)),
			  .by = c(age_group, lp)) |>
	arrange(age_group, lp) |> 
	gt(groupname_col = "age_group",
	   rowname_col = "lp") |> 
	cols_hide(lp) |> 
	tab_spanner("Vocabulary size ", matches("prop")) |> 
	cols_merge(c(total_prop_mean, total_prop_sd),
			   pattern = "{1} ({2})") |> 
	cols_merge(c(l1_prop_mean, l1_prop_sd),
			   pattern = "{1} ({2})") |> 
	cols_merge(c(l2_prop_mean, l2_prop_sd),
			   pattern = "{1} ({2})") |> 
	cols_merge(c(concept_prop_mean, concept_prop_sd),
			   pattern = "{1} ({2})") |> 
	cols_merge(c(te_prop_mean, te_prop_sd),
			   pattern = "{1} ({2})") |> 
	fmt_number(is.numeric,
			   decimals = 1,
			   scale_by = 100,
			   drop_trailing_zeros = FALSE) |>
	cols_label(total_prop_mean = "Total",
			   l1_prop_mean = "L1",
			   l2_prop_mean = "L2",
			   concept_prop_mean = "Concept",
			   te_prop_mean = "TE") |> 
	tab_stub_indent(rows = everything(),
					indent = 2) |> 
	tab_style(cell_text(align = "left"),
			  cells_stub())
```

@@

@@

We defined our time window of interest from 300 ms after the onset of the test phase (target and distractor presentation) until the end of the test phase (2,000 ms). For each trial, we chunked the time domain into 17 time bins of 100 ms of duration. We then calculated, for each experimental session, time bin, and condition, participant's proportion of target and distractor fixations. Finally, we computed the empirical logit of target fixations, which we introduced in the statistical analyses as our response variable. Missing eye-tracker samples were interpolated using the last-observation-carried-forward [see @zettersten2022peekbank for a similar approach].

We conducted two main analyses. First, we estimated the effect of phonological priming on participants' target looking, comparing *related* trials with *unrelated* trials. This analysis included all trials on the data set. Second, we estimated the effect of cognateness on phonological priming, comparing *cognate* with *non-cognate* trials, leaving out *unrelated* trials. In both analyses, we used General Additive Mixed Models (GAMMs) to model the probability of target fixations across the time course of the trial using a normal distribution.

In the first analysis. We included *Relatedness* (`Related` vs. `Unrelated`, sum-coded as `-0.5` and `+0.5`), *Group* (`Monolingual` vs. `Bilingual`, sum-coded as `-0.5` and `+0.5`), and *Age* (participants' standardised age in months) as fixed, main effects. We also included cubic regression splines for the main effect of *Time*, and one for an adjustment of the previous cubic spline by *Group* [@wood2017generalized]. For both splines, we specified $k = 10$ basis functions or *knots*--half the number of time bins, for computational convenience. Finally, we added by-participant random intercepts, and random slopes for the main effect of *Relatedness* and the main effect of *Age*, both including repeated measures per participant.

@@

$$
\begin{aligned}
\textbf{Likelihood:} \\
y_i &\sim \mathcal{N}(\mu_i, \sigma_i) \\ \\
\textbf{Linear model} \\
\text{logit}(\mu_i) &= (\beta_0 + u _{0_{i}}) + (\beta_1 + u _{1_{i}}) \cdot \text{Relatedness} + \beta_{2} \cdot \text{Group} + \\
&\beta_{3} \cdot (\text{Relatedness} \times \text{Group}) + (\beta_4 + u_{3_{i}}) \cdot \text{Age} + \\
&\sum_{j = 1}^k b_{j_{1}}(\beta_{5_{k}} + u_{4_{i}}) \cdot \text{Time} + \\
&\sum_{j = 1}^k b_{j_{1}} (\beta_{6_{k }} + u_{5_{i}}) \cdot (\text{Time} \times \text{Group}) \\
\text{where:} \\
&k \text{ is the number of knots in the spline (10)} \\
\textbf{Prior:} \\
\beta_{0-6} &\sim \mathcal{N}(0, 1) \\
b_{0-1} &\sim MVN(0, 1) \\
\sigma_i &\sim Exp(4) 
\end{aligned}
$$

@@

We know present the results under four different trial-level inclusion criteria. In Analysis 1, the child is required to understand the prime *and* the target words, on top of having to fixate both target and distractor  for at last 50 ms each during the target-distractor phase. In Analysis 2, the child is required to understand the target word, on top of having to fixate both target and distractor  for at last 50 ms each during the target-distractor phase. In Analysis 3, the child is not required to understand any word, but is still required to fixate both target and distractor for at last 50 ms each during the target-distractor phase. In Analysis 4, the child is not required to understand any word, nor to fixate at both target and distractor. In Analysis 5, the child is required to understand the prime *and* target words, but is not required to look at both target *and* distractor in the test phase.

::: {.callout-note}

@floccia2020translation required participants to know prime and target, but not to look at target *and* distractor (looking at at least one of them is enough). @mani2010infants did not require participants to know either word, but required infants to *not* look exclusively to the target or the distractor throughout the trial, and in an alternative analysis, to *not* look exclusively to the distractor.

The fact that in some time course analyses participants seem to be looking predominantly towards one of the objects (despite the target and distractor pictures having been presented less than 300 ms before) is a common finding in previous studies using preferential looking procedures [e.g., @floccia2020translation].

:::

>>>>

```{r setup}
#| label: setup
#| echo: false
#| message: false
#| warning: false
# load objects
targets::tar_config_set(store = here::here("_targets"),
						script = here::here("_targets.R"))

@@

targets::tar_load(
	c(participants,
	  stimuli,
	  vocabulary,
	  attrition_trials, 
	  attrition_participants,
	  bvq_data, 
	  gaze_aoi, 
	  data_related,
	  data_cognate,
	  data_time_related, 
	  data_time_cognate,
	))

targets::tar_load(
	c(model_fits_related,
	  model_fits_cognate,
	  # model_fits_related_time,
	  # model_fits_cognate_time,
	  # model_loo_related,
	  # model_loo_cognate,
	))

library(knitr)
library(kableExtra)
library(ggplot2)
library(gt)
library(patchwork)
library(english)
library(gtExtras)

# set ggplot theme and colour palette

@@

# prepare data
attrition <- lst(participants,
				 attrition_participants,
				 attrition_trials) |> 
	reduce(inner_join, by = join_by(session_id)) |> 
	unnest_wider(matches("all"), names_sep = "_")

@@

```{r}
#| label: participants-numbers
n_participants_total <- participants |> 
	distinct(location, child_id) |> 
	count(location)

n_participants_sex <- participants |> 
	distinct(location, sex, child_id) |> 
	count(location, sex)

n_participants_sessions <- count(participants,
								 child_id,
								 name = "n_sessions") |> 
	count(n_sessions) |> 
	group_split(n_sessions) |> 
	set_names(paste0("session_", 1:3))

@@

n_sessions_lp <- participants |> 
	summarise(across(age, lst(mean, sd, min, max)),
			  n = n(),
			  .by = c(lp)) |> 
	mutate(across(age_mean:age_max, \(x) round(x, 2))) |> 
	group_split(lp) |> 
	set_names(janitor::make_clean_names(unique(participants$lp)))

n_sessions_dominance_lp <- count(participants, lp, test_language) |> 
	group_split(test_language) |> 
	set_names(c("catalan", "english", "spanish")) |> 
	map(group_split, lp)

n_sessions_lp <- participants |> 
	count(lp) |> 
	group_split(lp) |> 
	set_names(c("monolingual_eng", "monolingual", "bilingual"))
```

We collected data from `r n_participants_total$n[1]` participants living in the Metropolitan Area of Barcelona (Spain) tested at the Laboratori de recerca en Infància at the Universitat Pompeu Fabra (`r n_participants_sex$n[1]` female, `n_participants_sex$n[2]` male), and `r n_participants_total$n[2]` participants living in the Oxfordshire area (United Kingdom), tested at the Oxford BabyLab at the University of Oxford (`r n_participants_sex$n[3]` female, `n_participants_sex$n[4]`, `n_participants_sex$n[5]` not reported). Families were recruited from maternity room in private hospitals or from social media, and contacted via phone when the child's age spanned between 20 and 32 months. From the `r sum(n_participants_total$n)` children that participated, `r n_participants_sessions$session_1$n` participated once, `r n_participants_sessions$session_2$n` participated twice, and `r n_participants_sessions$session_3$n` participated three times. Recurrent participants were tested with at least XXX months of difference. We gathered a total of `r n_sessions_total$n` testing sessions: `r n_sessions_lp$monolingual_english$n` from participants in Oxford, and `r n_sessions_lp$monolingual$n + n_sessions_lp$bilingual$n` from participants in Barcelona.

We assessed participants' language profile using the Language Exposure Questionnaire [LEQ, @bosch2001evidence]. Before each experimental session, the experimenter asked the caretakers to estimate the amount of hours per day they and other people in the infant's social circle have spent speaking to the infant in any language since birth. The output of this interview is an estimated degree of exposure (*DoE*) to each language, indicated by the proportion of time the infant was reported to have listened to each language. According to this estimate, we classified participants as Catalan-, English-, or Spanish-dominant if the language with highest *DoE* was Catalan, English, or Spanish, respectively, and tested the participant in the stimuli set that contained words in their native language. From the `r n_sessions_total$n` testing sessions, `r n_sessions_lp$monolingual_english$n` belonged to English monolinguals,  `r n_sessions_lp$monolingual$n` to Catalan or Spanish monolinguals, and  `r n_sessions_lp$bilingual$n` to Catalan-Spanish bilinguals. @tbl-participants shows a detailed description of the linguistic profile of our sample.

```{r tbl-participants}
#| label: tbl-partiicpants
#| tbl-cap: "Demographic and linguistic profile of participants."
participants |> 
	unnest_wider(doe) |> 
	select(-matches("others")) |> 
	mutate(lp = case_when(
		lp=="Monolingual" ~ "Monolingual (Catalan or Spanish)",
		lp=="Bilingual" ~ "Bilingual (Catalan-Spanish)",
		.default = lp)) |> 
	arrange(desc(lp)) |> 
	summarise(n = n(),
			  across(age, lst(mean, sd, min, max)),
			  across(matches("doe_"), lst(mean, sd)),
			  age_vctr = list(age),
			  .by = c(lp, test_language)) |> 
	gt(rowname_col = "test_language",
	   groupname_col = "lp") |> 
	gt_plt_dist(age_vctr, type = "histogram") |> 
	fmt_number(is.numeric, decimals = 1) |> 
	fmt_integer(n) |> 
	fmt_number(matches("doe"), decimals = 1, scale_by = 100) |> 
	cols_merge_range(age_min, age_max) |> 
	cols_merge_uncert(age_mean, age_sd) |> 
	cols_merge_uncert(doe_catalan_mean, doe_catalan_sd) |> 
	cols_merge_uncert(doe_english_mean, doe_english_sd) |> 
	cols_merge_uncert(doe_spanish_mean, doe_spanish_sd) |> 
	tab_spanner("Age (months)", matches("age")) |> 
	tab_spanner("Catalan", matches("catalan")) |> 
	tab_spanner("English", matches("english")) |> 
	tab_spanner("Spanish", matches("spanish")) |> 
	tab_spanner("Degree of Exposure (%)", matches("doe")) |> 
	cols_label(n = "N",
			   age_mean = "M (SD)",
			   age_min = "Range",
			   age_vctr = "Distribution",
			   doe_catalan_mean = "M (SD)",
			   doe_english_mean = "M (SD)",
			   doe_spanish_mean = "M (SD)") |> 
	tab_style(cell_text(style = "italic"),
			  cells_column_labels()) |> 
	tab_style(cell_text(weight = "bold"),
			  cells_column_spanners())
```

@@

```{r tbl-vocabulary}
#| label: tbl-vocabulary
#| tbl-cap: "Participant vocabulary sizes across ages and language profiles. Vocabulary size scores are presented using two measures: L1 vocabulary size (proportion of words marked as *Understands* in the vocabulary checklist of the dominant language), and total vocabulary size (proportion of words marked as *Understands* in vocabulary checklists of both dominant and non-dominant languages). Vocabulary size scores from the recurrent participants are shown linked."
#| eval: false
vocabulary |> 
	inner_join(select(participants, child_id, session_id, age, lp),
			   by = join_by(child_id, session_id)) |> 
	pivot_longer(c(total_prop, l1_prop),
				 names_to = "measure",
				 values_to = "prop",
				 names_transform = \(x) gsub("_prop", "", x)) |> 
	mutate(lp = case_when(
		lp=="Monolingual" ~ "Monolingual (Catalan or Spanish)",
		lp=="Bilingual" ~ "Bilingual (Catalan-Spanish)",
		.default = lp) |> 
			factor(levels = c("Monolingual (English)",
							  "Monolingual (Catalan or Spanish)",
							  "Bilingual (Catalan-Spanish)")),
		measure = if_else(measure=="l1", 
						  "L1 vocabulary", 
						  "Total vocabulary"),
		is_imputed = if_else(is_imputed, "Imputed", "Observed") |> 
			factor(levels = c("Observed", "Imputed"))) |> 
	ggplot(aes(age, prop, colour = is_imputed)) +
	facet_grid(measure ~ lp) +
	geom_line(aes(group = child_id),
			  alpha = 1/2,
			  colour = "grey") +
	geom_point() +
	labs(x = "Age (months)",
		 y = "Vocabulary size (%)") +
	theme(panel.grid = element_line(linetype = "dotted",
									colour = "grey"),
		  legend.position = "right",
		  legend.title = element_blank())
```

@@

## Data processing

```{r dataset}
n_trials <- nrow(attrition)
n_sessions <- n_distinct(attrition$session_id)
n_participants <- n_distinct(attrition$child_id)
n_trials_valid <- nrow(filter(attrition,
							  is_valid_participant,
							  is_valid_trial))
n_sessions_valid <- sum(attrition_participants$is_valid_participant)
n_participants_valid <- n_distinct(
	attrition[attrition$is_valid_participant, ]$child_id)

n_exc_prime <- sum(!attrition$is_valid_gaze_all_prime)
n_exc_test <- sum(!attrition$is_valid_gaze_all_test)
n_exc_test_each <- sum(!attrition$is_valid_gaze_all_test_each)
n_exc_vocab <- sum(!attrition$is_valid_vocab)
n_exc_participants <- sum(!attrition_participants$is_valid_participant)

n_longitudinal <- attrition |>
	filter(is_valid_participant,
		   is_valid_trial) |>
	distinct(child_id, session_n) |> 
	count(child_id, name = "times") |> 
	count(times)
```

Missing eye-tracker samples were interpolated using the last-observation-carried-forward [see @zettersten2022peekbank for a similar approach], with a maximum of 20 maximum consecutive missing samples being interpolated (an equivalent of 166.67 ms).

We gathered data from `r format(n_trials, big.mark = ",")` trials from `r n_sessions` testing sessions, generated from `r n_participants` distinct participants. We excluded trials in which participants failed to provide 50%  valid eye-tracking samples (equivalent to 750 ms) during the prime phase (*n* = `r format(n_exc_prime, big.mark = ",")`) or 50% valid samples (equivalent to 1,000 ms) during the target-distractor phase (*n* = `r format(n_exc_test, big.mark = ",")`). We also excluded trials in which participants did not provide at least 5% of valid samples (equivalent to 100 ms) of target-looking *and* distractor looking in the test phase (*n* = `r format(n_exc_test_each, big.mark = ",")`) [see @floccia2020translation for a similar approach]. Finally, we excluded trials in which participants did not understand the prime *or* the target word, according to a supplementary vocabulary checklist filled by their caregivers upon experiment completion (*n* = `r format(n_exc_vocab, big.mark = ",")`). After trials that matched any of the aforementioned exclusion criteria from the dataset, we excluded participants who did not provide at least two valid trials in the *Cognate* condition, the *Non-cognate* condition, or the  *Unrelated* condition (*n* = `r n_exc_participants`). The final dataset included `r format(n_trials_valid, big.mark = ",")` trials from `r n_sessions_valid` testing sessions, generated by `r format(n_participants_valid, big.mark = ",")` distinct participants. Of those participants, `r n_longitudinal[1, 2]` provided data from one experimental session, `r n_longitudinal[2, 2]` provided data from two experimental sessions, and `r n_longitudinal[3, 2]` provided data from three experimental sessions. @tbl-attrition-trials shows a detailed description of the trial attrition.

```{r tbl-attrition}
#| label: tbl-attrition
#| tbl-cap: "Participant-level and trial-level sample size after applying inclusion criteria. The number of excluded participants and trials is indicated between parentheses."
expanded <- expand(attrition, lp, trial_type, is_valid_trial)

n_lp <- attrition |> 
	distinct(is_valid_participant, lp, session_id) |> 
	count(is_valid_participant, lp) |> 
	pivot_wider(names_from = is_valid_participant,
				values_from = n,
				names_repair = janitor::make_clean_names) |> 
	mutate(n_participants_true = true,
		   n_participants_false = false) |> 
	select(lp, true, false)

attrition |> 
	summarise(n_trials_true = sum(is_valid_trial),
			  n_trials_false = sum(!is_valid_trial),
			  .by = c(lp, trial_type)) |> 
	arrange(desc(lp)) |> 
	# across(n_trials, lst(mean, sd, sum), na.rm = TRUE),

	pivot_wider(id_cols = lp,
				names_from = c(trial_type),
				values_from = matches("trials"),
				names_repair = janitor::make_clean_names) |> 
	left_join(n_lp, by = join_by(lp)) |> 
	mutate(lp = case_when(
		lp=="Monolingual" ~ "Monolingual (Catalan or Spanish)",
		lp=="Bilingual" ~ "Bilingual (Catalan-Spanish)",
		.default = lp)) |>
	select(lp, true, false,
		   matches("n_trials")) |> 
	gt(rowname_col = "lp") |> 
	tab_spanner("Participants", matches("participant")) |> 
	tab_spanner("Trials", matches("trial")) |> 
	fmt_integer(everything()) |> 
	cols_merge(c(true, false),
			   pattern = "{1} ({2})") |> 
	cols_merge(c(n_trials_true_cognate,
				 n_trials_false_cognate),
				 pattern = "{1} ({2})") |> 
	cols_merge(c(n_trials_true_non_cognate,
				 n_trials_false_non_cognate),
				 pattern = "{1} ({2})") |> 
	cols_merge(c(n_trials_true_unrelated, 
				 n_trials_false_unrelated),
				 pattern = "{1} ({2})") |> 
	cols_label(true = "N", 
			   n_trials_true_cognate = "Cognate",
			   n_trials_true_non_cognate = "Non-cognate",
			   n_trials_true_unrelated = "Unrelated") |> 
	tab_style(cell_text(style = "italic"),
			  cells_column_labels()) |>
	grand_summary_rows(fns = list(label = md("*N*")) ~ sum(.),
					   fmt = ~ fmt_integer(.))
```

@@

We conducted two main analyses. In the first analysis, we estimated the effect of phonological priming on participants' target looking comparing *related* trials---in which prime and target share phonological onset, and the prime was not a cognate word---with *unrelated* trials---in which prime and target did not share phonological onset, and neither was the prime a cognate word, leaving out *cognate* trials. This analysis included all trials on the data set. Second, we estimated the effect of cognateness on phonological priming, comparing *cognate* trials---in which prime and target shared phonological onset, and the prime was a cognate word that also shared phonological onset with the target, with *non-cognate* trials---in which prime a target shared phonological onset, and the prime was not a cognate---, leaving out *unrelated* trials.

In both analyses, we modelled the probability of target looking across the time course of the trial. For each trial, we chunked the time domain of the target-distractor phase into 20 time bins of 100 ms of duration. We then calculated, for each experimental session, time bin, and condition, the number of eye-tracker samples in which the participant was looking to the target and the distractor pictures. Finally, we computed the empirical logit of target fixations using @eq-elogit [@barr2008analyzing; @agresti2012categorical], which modelled in our analyses using a Gaussian distribution.

$$
\eta' = \ln\bigg(\frac{\text{Target} + 0.5}{ \text{Distractor} + 0.5}\bigg)
$$ {#eq-elogit}

In the first analysis, we included *Relatedness* (`Related` vs. `Unrelated`, sum-coded as `-0.5` and `+0.5`), *Group* (`Monolingual` vs. `Bilingual`, sum-coded as `-0.5` and `+0.5`), and *Age* (participants' standardised age in months) as fixed, main effects, together with their two- and three-way interactions. We used multilevel General Additive Mixed Models (GAMMs) to model the time course of target looking, including cubic regression splines for the main effect of *Time*, and one for an adjustment of the *Time* cubic spline by *Group* and *Relatedness* [@wood2017generalized]. For both splines, we specified $k = 10$ basis functions or *knots*--half the number of time bins---for computational convenience. Finally, we added by-participant random intercepts, and random slopes for the main effect of *Relatedness* and the main effect of *Age*, both including repeated measures per participant [@pedersen2019hierarchical]. In the second analysis, we included *Cognateness* (`Cognate` vs. `Non-cognate`, sum-coded as `-0.5` and `+0.5`) replacing *Relatedness*. @eq-model shows the formal implementation of the model.

@@

$$
\begin{aligned}
\textbf{Likelihood:} \\
y_i &\sim \mathcal{N}(\mu_i, \sigma_i) \\ \\
\textbf{Linear model:} \\
\eta'(\mu_i) &= (\beta_0 + u _{0_{i}}) + (\beta_1 + u _{1_{i}}) \cdot \text{[Condition]} +\\
&\beta_{2} \cdot \text{Group} + \beta_{3} \cdot (\text{[Condition]} \times \text{Group}) +\\
& (\beta_4 + u_{2_{i}}) \cdot \text{Age} + \sum_{j = 1}^k b_{j_{1}} \cdot \beta_{5_{k}} \cdot \text{Time} + \sum_{j = 1}^k b_{j_{1}} \cdot\beta_{6_{k}} \cdot (\text{Time} \times \text{Group}) \\
\text{where:} \\
&\eta'\text{ is the empirical logit of target fixations} \\
& \text{[Condition] should be replaced with } Relatedness \text{ or } Cognateness \\
&k \text{ is the number of knots in the spline (10)} \\
\textbf{Prior:} \\
\beta_{0-6} &\sim \mathcal{N}(0, 0.5) \\
u_{0-2} &\sim \mathcal{N}(0, \sigma_{0-2})\\
b_{0-1} &\sim \text{MVN}(0, \tau_{0-1}) \\
\sigma_{0-2}, \tau_{0-1} &\sim \text{Exponential}(6)\\
\rho_{0-2} &\sim LKJCorr(6)\\
\text{where:}\\
&\rho_{0-2} \text{ are the correlation parameters for } \sigma_{0-2}
\end{aligned}
$$ {#eq-model}

@@

```{r fig-ptlt-descriptive}
#| label: fig-ptlt-descriptive
#| fig-cap: "Average probability of target looking across the time course of the trial. Points an whiskers indicate the mean and standard error of the mean."
data_time_related |> 
	summarise(.elog = mean(.elog),
			  .by = c(session_id, timebin, lp, condition)) |> 
	ggplot(aes(timebin, .elog, colour = condition, fill = condition)) +
	facet_wrap(~lp) +
	geom_hline(yintercept = 0.5, linetype = "dashed") +
	stat_summary(fun.data = mean_se,
				 geom = "errorbar",
				 width = 0.5) +
	stat_summary(fun = mean,
				 geom = "line") +
	stat_summary(fun = mean,
				 geom = "point") +
	theme(axis.title.x = element_blank()) +

	data_time_cognate |> 
	summarise(.elog = mean(.elog),
			  .by = c(session_id, timebin, lp, condition)) |> 
	ggplot(aes(timebin, .elog, colour = condition, fill = condition)) +
	facet_wrap(~lp) +
	geom_hline(yintercept = 0.5, linetype = "dashed") +
	stat_summary(fun.data = mean_se,
				 geom = "errorbar",
				 width = 0.5) +
	stat_summary(fun = mean,
				 geom = "line") +
	stat_summary(fun = mean,
				 geom = "point") +
	theme(strip.text = element_blank()) +

	plot_layout(ncol = 1) &
	plot_annotation(tag_levels = "A") &
	labs(x = "Time bin (100 ms)",
		 y = "P(Target looking)",
		 colour = "Condition",
		 fill = "Condition") &
	# scale_y_continuous(breaks = seq(0, 1, 0.1)) &
	theme(legend.position = "right",
		  panel.grid.major.x = element_line(colour = "grey",
		  								  linetype = "dotted"),
		  panel.grid.major.y = element_line(colour = "grey",
		  								  linetype = "dotted"))
